---
title: "final-proj"
output: html_document
date: "2025-12-04"
---

This section of the project implements a Beta–Binomial Bayesian model to estimate, 
for each restaurant, the probability that a dish is unsafe for a given allergy. 
A restaurant level safety score (expected proportion of safe dishes) is calculated 
using the posterior distribution. Each restaurant is then classified as safe (1)
or unsafe (0) based on a threshold to make the decision making process even more
simple. 

```{r}
# import packages for visualization later
library(ggplot2)
library(knitr)

# dataset produced by first MLP step
dishes <- read.csv('~/Downloads/dishes_allergies.csv')

# function to run the model for a given allergy
run_bayes_for_allergy <- function(dishes,
                                  allergy,
                                  tau_i     = 0.90,
                                  alpha0     = 1,
                                  beta0      = 1,
                                  safe_thresh = 0.99,
                                  n_sims     = 5000) {
  
  # threshold to define unsafe dishes. introduced because the range of safety probability
  # scores was relatively narrow and the variability between restaurants were
  # small, so most of the restaurants sampled from the posterior were deemed "safe", 
  # making the model's results not very meaningful. 
  tau <- quantile(dishes[[allergy]], tau_i, na.rm = TRUE)
  
  # restaurant list
  restaurants <- unique(dishes$Restaurant)
  n_rest <- length(restaurants)
  
  # total dishes and unsafe dishes per restaurant
  n_items <- numeric(n_rest)
  k_items <- numeric(n_rest)
  
  # get all predicted allergy probabilities for each restaurant
  for (i in seq_len(n_rest)) {
    idx   <- dishes$Restaurant == restaurants[i]
    probs <- dishes[idx, allergy]
    
    n_items[i] <- length(probs) #  total dishes
    k_items[i] <- sum(probs > tau)   # unsafe dishes
  }
  
  # beta(1,1) prior -> posterior params
  alpha_post <- alpha0 + k_items
  beta_post  <- beta0  + (n_items - k_items)
  
  # posterior sampling
  theta_samples <- matrix(0, n_rest, n_sims)
  for (i in seq_len(n_rest)) {
    theta_samples[i, ] <- rbeta(n_sims, alpha_post[i], beta_post[i])
  }
  
   # E[unsafe rate]
  mean_theta   <- rowMeans(theta_samples)
  # expected fraction safe
  safety_score <- 1 - mean_theta 
  # deemed safe if score > safety threshold
  safe_binary  <- ifelse(safety_score > safe_thresh, 1, 0)
  
  # compile results into df
  results <- data.frame(
    Restaurant        = restaurants,
    Total_Dishes      = n_items,
    Unsafe_Dish_Count = k_items,
    Mean_Unsafe_Rate  = mean_theta,
    Safety_Score      = safety_score,
    Is_Safe           = safe_binary
  )
  
  # return everything in a list
  list(
    allergy  = allergy,
    tau          = tau,
    alpha0       = alpha0,
    beta0        = beta0,
    results      = results,
    safety_score = safety_score
  )
}
```


```{r}
# plug in lactose intolerance for test
allergy <- "prob_milk.allergy...lactose.intolerance"
bayes_out <- run_bayes_for_allergy(
  dishes = dishes,
  allergy = allergy
)
results       <- bayes_out$results
safety_score  <- bayes_out$safety_score
```

```{r}
# top 5 most safe
top_safe <- results[order(-results$Safety_Score), ][1:5, ]

# bottom 5 least safe 
bottom_5 <- results[order(results$Safety_Score), ][1:5, ]
bottom_5 <- bottom_5[order(-bottom_5$Safety_Score), ]

# format nicely
kable(top_safe,
      caption = paste("Top 5 Safest Restaurants for", allergy))
kable(bottom_5,
      caption = paste("Bottom 5 Least Safe Restaurants for", allergy))
```

```{r}
# plot distribution of safety scores across all restaurants
df <- data.frame(SafetyScore = safety_score)
N <- nrow(df)
binwidth <- 0.005

ggplot(df, aes(x = SafetyScore)) +
  geom_histogram(
    aes(y = ..count..),
    binwidth = binwidth,
    fill = "darkseagreen3",
    color = "darkgreen",
    alpha = 0.7
  ) +
  geom_density(
    aes(y = ..density.. * N * binwidth),
    size = 1,
    color = "darkgreen"
  ) +
  labs(
    title = "Distribution of Restaurant Safety Scores",
    x = "Safety Score",
    y = "Number of Restaurants"
  ) +
  theme_minimal(base_size = 10)
```


```{r}
# comparison of two methods (mlp and bayes ML)
mlp_scores <- read.csv("~/Downloads/mlp_results.csv")
names(mlp_scores)[2] <- "Safety_MLP"

bayes_results <- results[, c("Restaurant", "Safety_Score")]
names(bayes_results)[2] <- "Safety_Bayes"

combined <- merge(bayes_results, mlp_scores, by = "Restaurant")

# plot distributions against each other
dens_df <- rbind(
  data.frame(Safety = combined$Safety_MLP,  Method = "MLP"),
  data.frame(Safety = combined$Safety_Bayes, Method = "Bayesian")
)

ggplot(dens_df, aes(x = Safety, color = Method, fill = Method)) +
  geom_density(alpha = 0.25) +
  labs(
    title = "Distribution of Restaurant Safety Scores by Method",
    x = "Safety Score",
    y = "Density"
  ) +
  theme_minimal(base_size = 10)
```

```{r}
# plot comparison of restaurant safety scores for a milk allergy to see how well
# the two methods predict
ggplot(combined, aes(x = Safety_MLP, y = Safety_Bayes)) +
  geom_point(aes(color = abs(Safety_MLP - Safety_Bayes)),
             alpha = 0.6, size = 2) +
  scale_color_viridis_c(name = "Difference \n|MLP - Bayes|") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  coord_cartesian(xlim = c(0.6, 1.00),  
                  ylim = c(0.6, 1.00)) + 
  labs(
    title = "Comparison of MLP vs Bayesian Restaurant Safety Scores For Milk Allergy",
    x = "Safety Score (MLP)",
    y = "Safety Score (Bayesian)"
  ) +
  theme_minimal(base_size = 11)

```
